---
title: "R Programming"
output:
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: sentence
---

# R Programming

## 0.0 Data Types

We'll create variables using different R data types:

- **Numeric**
- **Character**
- **Logical**
- **Factor**
- **List**


### 0.1 Numeric
> The c() function is used to combine multiple values into a single vector.

> Purpose: To store quantitative values that can be used for arithmetic operations.
This line creates a numeric vector named age.
It contains three values: 25, 30, and 22, which represent ages.

```{r}
age <- c(25, 30, 22)
```

### 0.2 Character
> The values "Alice", "Bob", and "Carol" are strings (text)

```{r}
names <- c("Alice", "Bob", "Carol")
```


### 0.3 Logical
> This creates a logical (boolean) vector.
The values represent whether each person is a student or not.
TRUE and FALSE are R’s built-in logical constants.

```{r}
is_student <- c(TRUE, FALSE, TRUE)
```

### 0.4 Factor
> Factors are ideal for data analysis involving groups or categories (e.g., grades, gender, regions).
This creates a factor vector, which is used for categorical data.
> R treats factors as enumerated types, which can be ordered or unordered.
> You’ve explicitly set the levels (categories) from A to F, even though only A and B are used here.

```{r}
grades <- factor(c("A", "B", "A"), levels = c("A", "B", "C", "D", "F"))
```

### 0.5 Display types
```{r}
list_types <- list(
  Age = typeof(age),
  Names = typeof(names),
  StudentStatus = typeof(is_student),
  Grades = typeof(grades)
)

list_types
```

```{r}
students_df <- data.frame(
  Name = names,
  Age = age,
  Student = is_student,
  Grade = grades,
  stringsAsFactors = FALSE
)

```


### 0.6 Show the data frame
```{r}
knitr::kable(students_df, caption = "Student Information Data Frame")

info_list <- list(
  Names = names,
  Ages = age,
  Status = is_student
)
```


### 0.7 Loop through and print each component
```{r}
for (item in names(info_list)) {
  cat("\n🔹", item, ":\n")
  print(info_list[[item]])
}

students_df$Adult <- students_df$Age >= 25
```

### 0.7 Updated data frame
```{r}
knitr::kable(students_df, caption = "Updated Data Frame with 'Adult' Column")
```

## Data Analysis

> The `iris` dataset includes measurements of 150 iris flowers from three species.
>This report combines visualizations, and statistical methods for an end-to-end analysis.

## 1. Inspecting the Dataset

> First, let's get a feel for the data

### 1.1 Loading Dataset

> In R, there are many ways to load data, depending on the format of your data.
> Here we will load data from the dataset that come with R using the `library()` method

```{r}
library(datasets) 
```

### 1.2 View the first few rows of the dataset

> `head(iris)` will show you the first 6 rows, showing the column names and a sample of the data.

```{r}
head(iris)
```

### 1.3 Get the structure of the dataset (data types, number of observations/variables)

> `str()` show the type of data frame, number of observations (rows) and the number of the variables (columns).
> It will also show the data type of each column (e.g., num for numeric, Factor for species).

```{r}
str(iris)
```

### 1.4 Get the column names

> The `name()` displays the colunm headings for the dataset

```{r}
names(iris)
```

### 1.5 Get summary statistics for each column

> The `summary()` provides the minimum (min), 1st quartile, median, mean, 3rd quartile, and maximum (max) for the numerical columns, and the counts for each column (level of the Species factor).

```{r}
summary(iris)
```

### 1.6 Descriptive Statistics

> We examine summary statistics to understand the central tendency and dispersion of each numeric variable.
> Summary of numeric features to understand central tendency and spread.
> Loads the psych package, which provides more detailed descriptive statistics.

```{r}
summary(iris)
library(psych)
describe(iris[, 1:4])
```

### 1.7 Group-wise Summary by Species

> Compare mean feature values across species.

```{r}
aggregate(. ~ Species, data = iris[, 1:5], mean)
```

## 2. Visualisation

> To visualise the data we use the `plot()` method.

### 2.1 Creating a Basic Scatter Plot

> A scatter plot showing the relationship between Sepal Length and Petal Length, colored by species.

```{r}
plot(iris$Petal.Length)
```

### 2.1.1 Customising the Plot

> The `plot()` function has many arguments you can use to customize your graph.
> Here are some of the most common ones:

-   main: Sets the main title of the plot.
-   xlab: Sets the label for the x-axis.
-   ylab: Sets the label for the y-axis.
-   col: Changes the color of the points.
-   pch: Changes the plotting symbol (e.g., circles, triangles, squares).
-   cex: Changes the size of the plotting symbols.
-   type: Specifies the type of plot.
-   "p" for points (default)
-   "l" for lines
-   "b" for both points and lines
-   "o" for both points and lines, overlaid
-   "h" for histogram-like vertical lines
-   "s" for stair steps
-   "n" for no plotting (useful for setting up axes first)

```{r}
plot(iris$Sepal.Length, iris$Petal.Length,
     main = "Sepal vs Petal Length",
     xlab = "Sepal Length",
     ylab = "Petal Length",
     col = "blue",
     pch = 16
)

```

### 2.1.2 Further Customisation Plots

> To reveal how Petal Length varies with Sepal Length across species.
> Helpful in spotting clusters or patterns.
> Sepal Length (X) vs Petal Length (Y).
> Points are colored by Species (setosa, versicolor, virginica).
> A legend shows which color corresponds to which species.

```{r}
plot(iris$Sepal.Length, iris$Petal.Length,
     col = iris$Species,
     pch = 16,
     main = "Sepal vs Petal Length by Species",
     xlab = "Sepal Length",
     ylab = "Petal Length")
    legend("topleft", legend = levels(iris$Species), col = 1:3, pch = 16)
```

### 2.2 Box Plot

> Quickly compare the central tendency and spread of Petal Length by species.
> A boxplot displays the distribution of Petal Length across the three species, highlighting medians and variability.
> Our box plot shows the Petal Length by Species

```{r}
boxplot(Petal.Length ~ Species, data = iris, col = c("lightblue", "lightgreen", "lightpink"))

```

### 2.3 Histogram

> A histogram helps visualize the frequency distribution of Sepal Width for all flowers.
> Checks if Sepal Width is normally distributed or skewed, and to see the spread of values.
> Below is the Histogram of Sepal Width

```{r}
hist(iris$Sepal.Width, col = "skyblue", border = "white")
```

### 2.4 Pairwise Scatterplots (Scatterplot Matrix)

> This matrix of scatterplots shows pairwise relationships among all four numeric variables, with species color-coded

```{r}
pairs(iris[1:4], main = "Pairwise Plot", pch = 21, bg = c("red", "green3", "blue")[unclass(iris$Species)])
```

### 3.0 Cleaner Visualization

> A `ggplot2()` version of the scatter plot offering cleaner visualization and built-in legends.

```{r}
library(ggplot2)

ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point(size = 2) +
  labs(title = "Petal vs Sepal Length", x = "Sepal Length", y = "Petal Length") +
  theme_minimal()
```

### 3.1 Density Plot

> A density plot compares the distribution of Sepal Width across species using smooth curves.

```{r}
ggplot(iris, aes(x = Sepal.Width, fill = Species)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Sepal Width") +
  theme_minimal()
```

### 3.2 Interactive Visualization

> You can use the `plotly()` package:

```{r}
### install.packages("plotly")
library(plotly)
plot_ly(iris, x = ~Sepal.Length, y = ~Petal.Length,
        type = 'scatter', mode = 'markers',
        color = ~Species)
```

## 🔎 Statistical Analysis

### 4.0 Correlation Matrix

> We compute correlations between numeric variables to assess linear relationships.

```{r}
cor(iris[, 1:4])
```

### 4.1 Correlation Heatmap

```{r}
cor_matrix <- cor(iris[, 1:4])
heatmap(cor_matrix, col = colorRampPalette(c("white", "blue"))(20),
        main = "Correlation Matrix of Iris Features")
```

### 4.3 ANOVA

> ANOVA tests if the mean Petal Length is significantly different across the three species.
> We will use the `aov()` function

```{r}
anova_model <- aov(Petal.Length ~ Species, data = iris)
summary(anova_model)
```

### 4.4 Post-hoc Test (Tukey HSD)

> After an ANOVA test tells us that at least one group mean is different, it does not tell us which groups differ.
> This is where Tukey’s Honest Significant Difference (HSD) test `TukeyHSD()` comes in.

```{r}
anova_model <- aov(Petal.Length ~ Species, data = iris)
TukeyHSD(anova_model)
```

### 4.5 Linear Regression

> We fit a linear model using the `lm()` to predict Petal Length from Sepal Length and evaluate its explanatory power

```{r}
lm_model <- lm(Petal.Length ~ Sepal.Length, data = iris)
summary(lm_model)
```

### 4.6 Regression Diagnostic Plots

```{r}
plot(lm_model)
```

### 4.7 Principal Component Analysis (PCA)

> PCA function `prcomp()` reduces dimensionality while preserving as much variance as possible, helping us visualize data structure.

#### 4.7.1 PCA summary

```{r}
pca <- prcomp(iris[, 1:4], scale. = TRUE)
summary(pca)

```

#### 4.7.2 Scree Plot

> The scree plot shows the variance explained by each principal component.

```{r}
plot(pca, type = "l", main = "Scree Plot of PCA")
```

#### 4.7.3 PCA Biplot

> A biplot shows both the scores of each sample and the contribution of each variable.

```{r}
biplot(pca, col = c("gray", "red"))
```

#### 4.4 PCA Visualization with ggplot2

> A clearer visualization of PCA results using the first two principal components, colored by species.

```{r}
pca_df <- as.data.frame(pca$x)
pca_df$Species <- iris$Species

library(ggplot2)
ggplot(pca_df, aes(x = PC1, y = PC2, color = Species)) +
  geom_point(size = 2) +
  labs(title = "PCA of Iris Dataset", x = "PC1", y = "PC2") +
  theme_minimal()
```


## 5.0 Loading from External File

### 5.1 Installs pacman ("package manager") if needed
```{r}
if (!require("pacman")) install.packages("pacman")
```

### 5.2 Use pacman to load add-on packages as desired
```{r}
pacman::p_load(pacman, rio) 
```

### 5.3 Loading RIO 
```{r}
getwd()
install.packages("rio")
library(rio)
```

### 5.4 Importing CSV
```{r}
rio_csv <- import("mbb.csv")
head(rio_csv)
```

### 5.5 Importing TXT
```{r}
rio_txt <- import("mbb.txt")
head(rio_txt)
```

### 5.6 Excel XLSX
```{r}
rio_xlsx <- import("mbb.xlsx")
head(rio_xlsx)
```

### 5.7 Data Viewer
> This line opens the help page for the `View()` function in R.
It’s not needed in a report/output — it's for interactive help in the console.

> This opens the `rio_csv()` dataset in a spreadsheet-style viewer window (only works in interactive RStudio, not when rendering .Rmd).


```{r}
?View
View(rio_csv)
```

### 5.8 Read Table for TXT files 
> R's built-in function for text files (used by rio)

> `read.table()` Reads tabular data from a text file (default separator is any whitespace).
header = TRUE: Tells R the first row contains column names.

```{r}
r_txt1 <- read.table("mbb.txt", header = TRUE)

```
### 5.9 Customer Reader for TXT file
> `sep = "\t"` Specifies the tab character as the field separator — required for tab-separated files (.tsv or .txt with tabs).

> \t is for tabs, sep = "," for commas. R converts missing
 to "NA"
 
```{r}
r_txt2 <- read.table("mbb.txt", 
  header = TRUE, 
  sep = "\t")

```

### 5.10 Read CSV Files
>  R's built-in function for csv files (also used by rio)

> CSV files don't have to specify delimiters for missing data
> because CSV means "comma separated values"
> open using `read.csv()` fuction

```{r}
trends.csv <- read.csv("mbb.csv", header = TRUE)
```


## Key takeaways
> We combined visual and statistical tools to analyze the classic iris dataset.

- Petal Length differs significantly by species (ANOVA).

- Sepal Length is a strong predictor of Petal Length (Regression).

- PCA helps reduce dimensions while revealing structure across species.

- This workflow is a powerful template for analyzing other classification datasets.